---
title: "US Birth Statistics "
author: "Harald groven"
date: "10/14/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

[Description of assignement](https://docs.google.com/document/d/e/2PACX-1vQq5ozMyz6qoCpNOJNUOPQbyDQyNtVWcPEvJppYxp30YlOkHhUslTplCac8OJH80AN2PZnDtiUWl62W/pub)

# Download data frin cdc.gov

Since this is slow, I have commented out bash commands after runnding these first time. 

```{bash curl_dl, echo=TRUE}
# Download zip file to working directory 

# curl -O ftp://ftp.cdc.gov/pub/Health_Statistics/NCHS/Datasets/DVS/natality/Nat2017us.zip
# unzip Nat2017us.zip
```

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(readr)

# less awful HTML table design
# devtools::install_github("haozhu233/kableExtra")
library(knitr)
library(kableExtra)
```



Census microdata for all 2017 births in the US is concained. Each row reprsents one birth. 
Fixed file format, i.e. relative position of each variable are always the same for each line. 

In a sane world, the data would be provided in a format where the codebook is in another format than PDF. The most efficient way of solving this exercise would have been to download the data as a STATA, SAS or SPSS files from [National Bureau of Economic Research](http://www.nber.org/data/vital-statistics-natality-data.html), and convert the data to R. 

The fixed format datafile has the name Nat2017PublicUS.c20180516.r20180808.txt when unzipped. The R function read_fwf is able to parse this file format. 

Variable name  | Position in fwf file 
------------- | -------------
dob_mm  | From position 13 to position 14 
sex  | From position 475 to position 475 
dbwt  | From position 504 to position 507 

 
```{r}

census_file <- "Nat2017PublicUS.c20180516.r20180808.txt"

# 4. Named arguments with start and end positions
us_census <- read_fwf(census_file, 
                      guess_max = 1000,
                      fwf_cols(
                        dob_mm = c(13, 14), 
                        sex = c(475, 475), 
                        dbwt = c(504, 507)
                        ), 
                      col_types = "ici"
                      )
# View(us_census)

```


# What is the proportion of boys to girls?

```{r bsex}
# group by variable sex. Store total number of observations in variable n_births. 
us_census %>% 
  group_by(sex) %>%  
  summarise (n_births = n()) %>%  
  mutate(freq = 100*n_births / sum(n_births)) %>% 
  kable() %>%
  kable_styling()
```

# What is the average birth weight in grams?

Calculating mean value on variable dbwt gives the following result: 

```{r bweight}
mean(us_census$dbwt)%>% 
  kable() %>%
  kable_styling()
```

# Make a density plot on the birth weight between boys and girls. Is it different?

First filter out unknown values, which is coded as 9999. 
A density plot (of type geom_density) yield the following result. 

```{r}
# thanks to http://www.sthda.com/english/wiki/ggplot2-density-plot-quick-start-guide-r-software-and-data-visualization for tips 

us_census %>% 
  filter(dbwt!="9999") %>%
  ggplot(., aes(x=dbwt, group=sex, fill=sex))+
    geom_density(alpha=0.15)+
    scale_fill_manual(values=c("#ff0000", "#0000ff"))+
    xlab("Birth weight")


```


# Does the average birth weight vary by month?

Birth weight seems pretty consistent for all months between 3250 and 3270 gram. However, birth weight is sligtly lower in December and January. 

```{r warning=FALSE}
us_census %>% 
  filter(dbwt!="9999") %>%
  select(dob_mm, dbwt) %>% 
  group_by(dob_mm) %>% 
  summarise(mean_birth_weight = mean(dbwt))%>% 
  kable() %>%
  kable_styling()
```


```{r warning=FALSE}
us_census %>% 
  filter(dbwt!="9999") %>%
  select(dob_mm, dbwt) %>% 
  group_by(dob_mm) %>% 
  summarise(mean_birth_weight = mean(dbwt))%>% 
  ggplot(., aes(x=dob_mm, y=mean_birth_weight))+
    geom_bar(stat = "summary")+
    coord_cartesian(ylim=c(3250,3270))+
    # show months as inteers, not decimal numbers 
    scale_x_continuous(breaks=c(1:12))+
    ylab("Mean birth weight")+
    xlab("Month number")+
    ggtitle("Mean birth weight by month")
```