---
title: "oblig6"
author: "Harald Groven"
date: "10/29/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# load packages
library(tidyverse)
# string functions 
library(stringr)
# graphics 
library(ggplot2)
# table styles 
library(kableExtra)
```


## Create tribble `DS` with 6 variables 

```{r echo=TRUE}
DS = tribble(
 ~fk_account_code, ~Ansvar, ~fk_function_code, ~fk_project_code, ~amount,
 101030, 40220, "Det", 3432, 1493.00,
 101030, 40220, "Met", 3586, 2827.00,
 101030, 40320, "Det", 3456, 49440.00,
 101030, 40330, "Sal", NA, 870716.00,
 101030, 40350, "Met", NA, 559928.00,
 101030, 40360, "Sal", NA, 125534.00,
 101030, 40280, "Pol", NA, 251611.00
 )
# View(DS)
```

## Drop a variable

1) Remove the `fk_project_code` variable from `DS`.

In tidyr this may be achieved by selecting the data frame minus the unwanted columnt. This has the same effect as dropping the variable. 

```{r echo=TRUE}
DS <- select(DS, -fk_project_code)
```

2) Sum the `amount` by the 3 first digits of `Ansvar`; 402, 403, etc.

Created new variable `ansvar3` containing the 3 first digits of Ansvar by using the regular expression `\\d{3,3}`
This new variable could then be used for grouping and summarising. 

```{r echo=TRUE}

DS %>% 
  mutate(ansvar3 = str_extract(DS$Ansvar, "\\d{3,3}")) %>% 
  select(ansvar3, amount) %>% 
  group_by(ansvar3) %>% 
  summarise(sum = sum(amount)) %>% 
  # style   
  kable() %>%
  kable_styling()
```

## Recode values 

3) Make new "labels" for `fk_function_code`, where:
        `Det` and `Sal` is `supplies`,
        `Met` is `inventories` and
        `Pol` is `other expenses`
        
Created a new variable `fk_function_label` as a copy of `fk_function_code`. Then used the `recode` function to replace codes with readable labels. 

```{r echo=TRUE}

DS <- DS %>%
  mutate(fk_function_label = recode(DS$fk_function_code, 
       "Det" = "supplies", 
       "Sal" = "supplies",
       "Met" = "inventories", 
       "Pol" = "other expenses"
       ))
# View(DS)
```



## Sum and group by 

Calculate the share of sales per product per year. The sum over the 3 shares per year is 100.

```{r echo=TRUE}
# first read some toy data 
df <- data.frame(Product=gl(3,10,labels=c("A", "B", "C")),
                Year=factor(rep(2002:2011, 3)),
                Sales=1:30)
# View(df)

df <- df %>% 
  group_by(Year) %>% 
  mutate(sum_year = sum(Sales)) %>% 
  ungroup() %>%
  mutate(pst_sum_year = (100/sum_year)*Sales) 

df %>% 
  # style   
  kable() %>%
  kable_styling()

```

## Stacked bar chart

Make a plot of the sales and shares per year per company.

Easiest possible stacked bar chart may be created with `ggplot2`'s `geom_bar`. Variable `Product` used for stacking within year. 

```{r echo=TRUE}

ggplot(df) + 
  geom_bar(aes(x=Year, y=Sales, fill = Product), stat="identity")

```

